Controlled Experiments for Word Embeddings

### 2015-09-09 article/main.tex
final draft

 \author{
 	Benjamin Wilson\\
	Lateral GmbH\\
	\texttt{benjamin@lateral.io}
	\And
	Adriaan Schakel\\
	NNLP\\
	\texttt{adriaan.schakel@gmail.com}
 }

@@

\begin{section}{Introduction}
Word embeddings, or distributed representations of words, have been the subject
of much recent research in the natural language processing and machine learning
communities, demonstrating state-of-the-art performance on word similarity and
word analogy tasks, amongst others.
Word embeddings represent words from the vocabulary as dense, real-valued
vectors.  Instead of one-hot vectors that merely indicate the location
of a word in the vocabulary, dense
vectors of dimension much smaller than the vocabulary size are
constructed such that they carry syntactic and semantic information.
Irrespective of the technique chosen, word embeddings are
typically derived from word co-occurrences.  More specifically, in a
machine-learning setting, word embeddings are typically trained by
scanning a short window over all the text in a corpus.  This process can
be seen as sampling word co-occurrence distributions, where it is
recalled that the co-occurrence distribution of a target word $w$
denotes the conditional probability $\p(w'|w)$ that a word $w'$ occurs in
its context, i.e. given that $w$ occurred.  Most applications of word
embeddings explore not the word vectors themselves, but relations
between them to solve, for example, similarity and word relation tasks
\cite{vecchi-baroni-zamparelli2011}.  For these tasks, it was found that using normalised word
vectors improves performance.  Word vector length is therefore typically
ignored.

@@

This paper is structured as follows.  Section \ref{related-work} draws
connections to related work, while Section \ref{corpus-and-model} describes
the corpus and the CBOW model used in our experiments.
Section \ref{WFVE} describes a controlled experiment for varying word
frequency while holding the co-occurrence distribution fixed.
Section \ref{CNVE}, in a complementary fashion, describes a controlled
experiment for varying the level of noise in the co-occurrence
distribution of a word while holding the word frequency fixed.  The
final section, Section \ref{future-directions}, considers further questions
and possible future directions.
\end{section}

\begin{section}{Related work}\label{related-work}
Our notion of word significance is reminiscent of the concept of
``contextual distinctiveness'' introduced earlier by McDonald and
Shillcock \cite{mcdonald2001contextual}.
They define the contextual distinctiveness of a target word $w$ as the
Kullback-Leibler divergence
between the unconditional word distribution $\p(w')$, and the
co-occurrence distribution $\p(w' | w)$ of the target word $w$.
So defined, contextual distinctiveness is a measure of the amount of
information provided by word $w$ about its contexts of use.  More specifically, if a word $w$ tends to appear in a wide variety of contexts, then its co-occurrence distribution tends to diverge less from unconditional word distribution, and its
contextual distinctiveness is accordingly small.  If, on the other hand,
$w$ typically appears in only a small number of different contexts, its co-occurrence distribution tends to diverge more from the unconditional word distribution, and a larger value of contextual distinctiveness results.
In order to reduce the computational burden, the authors considered only $690$ target words, randomly chosen from equally divided frequency bands.
Furthermore, the authors considered only co-occurrence with the $500$ most frequent content words.
This significant restriction, necessary in their case in order to reduce the variance of the estimates of their measure of contextual distinctiveness, does not apply in our study.

Our experimental finding that word vector length decreases with
co-occurrence noise is related to earlier work by Vecchi, Baroni, and
Zamparelli \cite{vecchi-baroni-zamparelli2011}, where a relation between
vector length and the ``semantic deviance'' of an adjective-noun
composite was studied empirically.  
In that paper, which is also based
on word co-occurrence, the authors study adjective-noun composites.
They built a vocabulary from the 8k most frequent nouns and 4k most frequent adjectives in a large general language corpus and added 22k adjective-noun composites.
For each item in the vocabulary, they recorded the co-occurrences with the top 10k most frequent content words (nouns, adjectives or verbs), and contructed word embeddings via singular value decomposition of the co-occurrence matrix \cite{landauer-dumais1997}.
The authors considered several models for constructing vectors of unattested adjective-noun composites, the two simplest being adding and component-wise multiplying the adjective and noun vectors.
They hypothesized that the length of the vectors thus constructed can be used to distinguish acceptable and semantically deviant adjective-noun composites.
Using a few hundred adjective-noun composites selected by humans for evaluation, they found that deviant composites have a shorter vector than acceptable ones, in accordance with their expectation.
In contrast to their work, our approach does not require extensive human evaluation.

Recent theoretical work \cite{Arora2015} has approached the problem of explaining the so-called ``compositionality'' property exhibited by some word embeddings.
In that work, unnormalised vectors are used in their model of the word relationship task.
It is hoped that experimental approaches such as those described here might enable theoretical investigations to describe the role of the word vector length in the word relationship tasks.
\end{section}

\begin{section}{Corpus and model}\label{corpus-and-model}
Our training data is built from the Wikipedia data dump from October
2013.  To remove the bulk of robot-generated pages from the training
data, only pages with at least 20 monthly page views are
retained.\footnote{For further justification and to obtain the dataset,
  see\\ \blogpost} Stubs and disambiguation pages are also removed,
leaving 463 thousand pages with a total of 482 million words.
Punctuation marks and numbers were removed from the pages and all words
were lower-cased.  Word frequencies are summarised 
in Table \ref{word-occurrences-table}.  This base corpus is then
modified as described in Sections \ref{WFVE} and \ref{CNVE}.  For
recognisability, the tokens inserted into the corpus are upper-cased.

\begin{subsection}{Word2vec}\label{word2vec}
Word2vec, a feed-forward neural network with a single hidden
layer, learns word vectors from word co-occurrences in an unsupervised
manner.  Word2vec comes in two versions.  In the continuous bag-of-words
(CBOW) model, the words appearing around a target word serve as input.
That input is projected linearly onto the hidden layer and the networks
then attempts to predict the target word on output.  Training is
achieved through back-propagation.  The word vectors are encoded in the
weights of the first synaptic layer, ``syn0''.  The weights of the
second synaptic layer (``syn1neg'', in the case of negative sampling) are typically discarded.  In the other model,
called skip-gram, target and context words swap places, so that the
target word now serves as input, while the network attempts to predict
the context words on output.

@@

The geometric distribution is truncated in order to limit the number of
tokens inserted into the corpus.  For any choice $0 < p < 1$ and
maximum value $n > 0$, the truncated geometric distribution is given by
the probability density function
\begin{equation} 
\label{distro}
P_{p, n} (i) = \frac{p^{i-1}  (1-p)}{1 - p^n}, \qquad 1
\leqslant i \leqslant n.
\end{equation} 
The factor in the denominator, which tends to unity in the limit $n \to
\infty$, assures proper normalisation.  We have chosen this distribution
because the probabilities decay exponentially base $p$ as a function of
$i$.  Of course, other distributions might equally well have been chosen
for the experiments.
\end{subsection}
\end{section}

@@

\begin{subsection}{Tokens derived from existing words}\label{WFVEexisting}
We choose uniformly at random a small number of words from the
unmodified vocabulary for our experiment.  In order that the inserted
tokens do not have too low a frequency, only words which occur at least
10 thousand times are chosen.  We also include the high-frequency
stopword \word{the} for comparison.  Table \ref{word-frequency-counts}
lists the words chosen for this experiment along with their frequencies.

The replacement procedure of Section \ref{replacement-procedure} is then
performed for each of these words, using a geometric decay rate of $p =
\tfrac{1}{2}$, and maximum value $n=20$, so that the 1st token is
inserted with a probability of about $0.5$, the 2nd with a probability of
about $0.25$, and so on.  This value of $p$ is one of a range of values
that ensure that, for each word, multiple tokens will be inserted with a
frequency sufficient to survive the low-frequency cut-off of
128.  A maximum value $n=20$ suffices for this choice of $p$, since
$2^{20 + \log_2{128}}$ exceeds the maximum frequency of any token in
the corpus.  Figure \ref{fig:word-frequency-experiment-text-cat}
illustrates the effect of these modifications on a sample text, with a
family of tokens \word{CAT\_i}, derived from the word \word{cat}.
Notice that all occurrences of the word \word{cat} have been replaced
with the tokens \word{CAT\_i}.

\begin{table}
	\input{word-frequency-experiment-counts.tex}
	\caption{Words chosen for the word frequency variation experiment, along with their frequency in the unmodified corpus. }
	\label{word-frequency-counts}
\end{table}

\begin{figure}
	\begin{mdframed}
	\input{word-frequency-experiment-text-cat.tex}
	\end{mdframed}
	\caption{Example sentences modified in the word frequency
          experiment as per Section \ref{WFVEexisting}, where the word
          \word{cat} is replaced with tokens \word{CAT\_i} using the
          truncated geometric distribution (\ref{distro}) with
          $p=\tfrac{1}{2}$ and $n=20$.}
	\label{fig:word-frequency-experiment-text-cat}
\end{figure}

@@

\begin{subsection}{Tokens derived from an artificial, meaningless word}\label{WFVEmeaningless}
Whereas the tokens introduced above all replace an existing word that
carries a meaning, we now include for comparison a high-frequency,
meaningless word.  We choose to introduce an artificial, entirely
meaningless word \word{VOID} into the corpus, rather than choose an
existing (stop)word whose meaninglessness is only supposed.  To achieve
this, we intersperse the word uniformly at random throughout the corpus so
that its relative frequency is $0.005$.  The co-occurrence
distribution of \word{VOID} thus coincides with the unconditional word
distribution.  The replacement procedure is then performed for this
word, using the same values for $p$ and $n$ as above.
Figure \ref{fig:word-frequency-experiment-text-void} shows the effect of
these modifications on a sample text, where a higher relative frequency of $0.05$ is used instead
for illustrative purposes.  Although the probability of inserting token
\word{VOID\_1} is twice that of inserting token \word{VOID\_2} and four
times that of inserting token \word{VOID\_3}, these three tokens all
appear once in this small piece of text.

\begin{figure}
	\begin{mdframed}
	\input{word-frequency-experiment-text-void.tex}
	\end{mdframed}
	\caption{The same example sentences as in
          Figure \ref{fig:word-frequency-experiment-text-cat} where
          instead of the word \word{cat} now the meaningless word
          \word{VOID} is replaced with tokens \word{VOID\_i}.  For
          illustrative purposes, the meaningless word \word{VOID} was
          here interspersed with a relative frequency of $0.05$.}
	\label{fig:word-frequency-experiment-text-void}
\end{figure}
\end{subsection}

@@

\subsubsection{Word frequency and vector direction}\label{WFVE-direction}
Figure \ref{word-frequency-experiment-heatmap} shows the cosine
similarity of pairs of vectors representing some of the tokens used in
this experiment.  Recall that the cosine similarity measures the extent
to which two vectors have the same direction, taking a maximum value of
$1$ and a minimum value of $-1$.  The number of different tokens
associated with an experiment word is the number of times that its
frequency can be halved and remain above the low-frequency cut-off of
$128$.

Consider first the words other than \word{the} and \word{VOID}.  These
words typically occur in a limited number of distinct contexts, and
so a limited number of samples is required to capture their
co-occurrence distributions.  We therefore expect that the vectors
representing the tokens derived from any of these words can be properly
learned from the modified corpus, even though their combined frequency equals the frequency of that word in the original
corpus.  The red blocks in Figure \ref{word-frequency-experiment-heatmap}
clearly demonstrate that the direction of these vectors is constant for
each family of tokens.  In other words, the direction of these vectors
is only determined by the original word and is independent of how often
that words appear, provided sufficient samples have been processed for
the word vector to settle in its equilibrium direction.

The artificial word \word{VOID}, on the other hand, occurs by
construction in every context.  A much higher number of samples is
therefore required to capture its co-occurrence distribution, and
thereby to learn its word vector.  The same is true, but to a lesser
extent, for the stopword \word{the}.  The difficulty of learning the
vector representing \word{VOID\_i} or \word{THE\_i} increases with $i$,
for the tokens occur increasingly less often.  This is consistent with
Figure \ref{word-frequency-experiment-heatmap}, where it is apparent that
the vectors representing \word{VOID\_i} for small $i$, say $i = 1,
\dots, 5$, roughly share a common direction, while this is no longer
true for higher values of $i$.  The same trend is apparent, although
less extremely so, for the word \word{the}.  The vectors for the tokens
\word{THE\_i} share a common direction for values of $i$ as high as
$14$, before they start changing direction.  We conclude that these
deviations are artifacts of poor learning due to an insufficient number of occurrences of these tokens.
Table \ref{word-frequency-experiment-cosine-similarity}, showing the
most similar words to each token \word{THE\_i}, supports this further.  In the
discussions below, the tokens \word{VOID\_i} and
\word{THE\_i} whose vectors have a cosine similarity less than $0.8$
with \word{VOID\_1} and \word{THE\_1}, respectively are therefore
discarded.

@@

\subsubsection{Word frequency and vector length}
We next consider the effect of frequency on word vector length.
Figure \ref{fig:word-frequency-experiment-graph} shows this relation for
individual words, both for the word vectors, represented by the weights
of the first synaptic layer, syn0, in the word2vec neural network, and
for the vectors represented by the weights of the second synaptic layer,
syn1neg.  We include the latter, which are typically ignored, for
completeness.  Each line corresponds to a single word, and the points on
each line indicate the frequency and vector length of the tokens derived
from that word.  For example, the six points on the line corresponding
to the word \word{protestant} are labeled, from right to left, by the
tokens \word{PROTESTANT\_1}, \word{PROTESTANT\_2}, \dots,
\word{PROTESTANT\_6}.  Again, the number of points on the line is
determined by the frequency of the original word.  For example, the
frequency of the word \word{protestant} can be halved at most $6$ times
so that the frequency of the last token is still above the
low-frequency cut-off.  Because all the points on a line share the same
co-occurrence distribution, the left panel in
Figure \ref{fig:word-frequency-experiment-graph} demonstrates conclusively
that length does indeed depend on frequency directly.  Moreover, this
relation is seen to be approximately linear for each word considered.
Notice also that the relative positions of the lengths of the word vectors associated
with the experiment words are roughly independent of the frequency band,
i.e. the plotted lines rarely cross.

Notice the inflection of the syn1neg curve in
Figure \ref{fig:word-frequency-experiment-graph} for \word{the} at
approximately $10^6$.  We have no explanation of this.  Also observe
that the lengths of the vectors representing the meaningless tokens
\word{VOID\_i} are approximately constant.  Since we already found the
direction to be also constant, it is sensible to speak of the word
vector of \word{VOID} irrespective of its frequency.  In particular, the
token \word{VOID\_1} may be taken as an approximation.

@@

\begin{sidewaysfigure*}
	\centering{\includegraphics[scale=0.6]{word-frequency-experiment-graph}}
	\caption{ Vector length vs. frequency for tokens derived from a
          few words chosen at random.  For each word, tokens of varying
          frequency but with the co-occurrence distribution of that word
          were inserted into the corpus, as described in
          Section \ref{WFVE}.  The vectors are obtained from the first
          synaptic layer, syn0, of the word2vec neural network.  The
          vectors obtained from the second layer, syn1neg, are
          included for completeness.
	  Legend entries are ordered by the L2-norm at the lowest-frequency
  (i.e. left-most) sample point for the corresponding syn0 plot, descending.}
	\label{fig:word-frequency-experiment-graph}
\end{sidewaysfigure*}
\end{section}

\begin{section}{Varying co-occurrence noise}\label{CNVE}
This second experiment is complementary to the first.  Whereas in the
first experiment we studied the effect of word frequency on word vectors
for fixed co-occurrence, we next study the effect of co-occurrence noise when the frequency is fixed.
As before, we do so in a controlled manner.

\subsection{Generating noise}
We take the noise distribution to be the (observed) unconditional word distribution.
Thus noise can be added to the co-occurrence distribution of a word by simply interspersing occurrences of
that word uniformly at random throughout the corpus.  A word that is
consistently used in a distinctive context in the unmodified corpus thus
appears in the modified corpus also in completely unrelated contexts.
As in Section \ref{WFVE}, we choose a small number of words from the
unmodified corpus for this experiment.  Table
\ref{cooccurrence-noise-words} lists the words chosen, along with their
frequencies in the corpus.

For each of these words, the replacement procedure of Subsection
\ref{replacement-procedure} is performed using a geometric rate of decay
of $p = \tfrac{5}{6}$, and truncating the distribution at $n=8$.  For
every replacement token (e.g. \word{CAT\_i}), occurrences of this
token are subsequently interspersed uniformly at random throughout the
corpus, such that the frequency of the replacement token is restored to
that of the original word \word{cat}.  For example, if the original word
\word{cat} occurred $1000$ times, then after the replacement procedure,
\word{CAT\_2} occurs approximately $694$ times, so a further
(approximately) $306$ random occurrences of \word{CAT\_2} are
interspersed throughout the corpus.  Thus the word \word{cat} is removed
from the corpus, and a family of tokens \word{CAT\_i}, $1 \leqslant i
\leqslant 8$ is introduced.  These tokens all have the same frequency as
the word \word{cat}, but their co-occurrence distributions, while based
on that of \word{cat}, have an increasing amount of noise.

Figure \ref{fig:co-occurrence-noise-experiment-text} illustrates the
effect of this modification in the case where the only word chosen is
\word{cat}.  The original text in this case concerned both cats and
dogs.  Notice that the word \word{cat} has been replaced entirely in the
cats section by \word{CAT\_i} and, moreover, that these same tokens
appear also in the dogs section. These occurrences (and additionally,
with probability, some occurrences from the cats section) constitute
noise.

The proportion of occurrences of the token \word{CAT\_i} that arise by
replacing the word \word{cat} is $P_{5/6, 8}(i)$, and the proportion
that arise from randomly interspersing this token throughout the corpus
is $r_i := 1 - P_{5/6, 8}(i)$.  The particular value of $p$ (large
compared to that of Section \ref{WFVE}) was chosen so that the
proportion of noise $r_i$ did not increase too quickly with $i$.  These
proportions are the sampling points on our axis of variation, and this
value of $p$ (together with the choice of $n$) assures a reasonable
spread.  Other parameter values (or indeed other distributions) could,
of course, have been used equally well.

>>>>

 \author{
 	Benjamin Wilson\\
	Lateral GmbH\\
	\texttt{benjamin@lateral.io}
	\And
	Adriaan M.\,J.\, Schakel\\
	NNLP\\
	\texttt{adriaan.schakel@gmail.com}
 }

@@

\begin{section}{Introduction}
Word embeddings, or distributed representations of words, have been the
subject of much recent research in the natural language processing and
machine learning communities, demonstrating state-of-the-art performance
on word similarity and word analogy tasks, amongst others.  Word
embeddings represent words from the vocabulary as dense, real-valued
vectors.  Instead of one-hot vectors that merely indicate the location
of a word in the vocabulary, dense vectors of dimension much smaller
than the vocabulary size are constructed such that they carry syntactic
and semantic information.  Irrespective of the technique chosen, word
embeddings are typically derived from word co-occurrences.  More
specifically, in a machine-learning setting, word embeddings are
typically trained by scanning a short window over all the text in a
corpus.  This process can be seen as sampling word co-occurrence
distributions, where it is recalled that the co-occurrence distribution
of a target word $w$ denotes the conditional probability $\p(w'|w)$ that
a word $w'$ occurs in its context, i.e., given that $w$ occurred.  Most
applications of word embeddings explore not the word vectors themselves,
but relations between them to solve, for example, similarity and word
relation tasks \cite{dontcountpredict}.  For these tasks, it
was found that using normalised word vectors improves performance.  Word
vector length is therefore typically ignored.

@@

This paper is structured as follows.  Section~\ref{related-work} draws
connections to related work, while Section~\ref{corpus-and-model} describes
the corpus and the CBOW model used in our experiments.
Section~\ref{WFVE} describes a controlled experiment for varying word
frequency while holding the co-occurrence distribution fixed.
Section~\ref{CNVE}, in a complementary fashion, describes a controlled
experiment for varying the level of noise in the co-occurrence
distribution of a word while holding the word frequency fixed.  The
final section, Section~\ref{future-directions}, considers further questions
and possible future directions.
\end{section}

\begin{section}{Related work}\label{related-work}
Our notion of word significance is reminiscent of the concept of
``contextual distinctiveness'' introduced earlier by McDonald and
Shillcock \cite{mcdonald2001contextual}.  They define the contextual
distinctiveness of a target word $w$ as the Kullback-Leibler divergence
between the unconditional word distribution $\p(w')$, and the
co-occurrence distribution $\p(w' | w)$ of the target word $w$.  So
defined, contextual distinctiveness is a measure of the amount of
information provided by word $w$ about its contexts of use.  More
specifically, if a word $w$ tends to appear in a wide variety of
contexts, then its co-occurrence distribution tends to diverge less from
unconditional word distribution, and its contextual distinctiveness is
accordingly small.  If, on the other hand, $w$ typically appears in only
a small number of different contexts, its co-occurrence distribution
tends to diverge more from the unconditional word distribution, and a
larger value of contextual distinctiveness results.  To reduce the
computational burden, the authors considered only $690$ target words,
randomly chosen from equally divided frequency bands.  Furthermore, the
authors considered only co-occurrence with the $500$ most frequent
content words.  This significant restriction, necessary in their case in
order to reduce the variance of the estimates of their measure of
contextual distinctiveness, does not apply in our study.

Our experimental finding that word vector length decreases with
co-occurrence noise is related to earlier work by Vecchi, Baroni, and
Zamparelli \cite{vecchi-baroni-zamparelli2011}, where a relation between
vector length and the ``semantic deviance'' of an adjective-noun
composite was studied empirically.  In that paper, which is also based
on word co-occurrence statistics, the authors study adjective-noun
composites.  They built a vocabulary from the 8k most frequent nouns and
4k most frequent adjectives in a large general language corpus and added
22k adjective-noun composites.  For each item in the vocabulary, they
recorded the co-occurrences with the top 10k most frequent content words
(nouns, adjectives or verbs), and constructed word embeddings via
singular value decomposition of the co-occurrence matrix
\cite{landauer-dumais1997}.  The authors considered several models for
constructing vectors of unattested adjective-noun composites, the two
simplest being adding and component-wise multiplying the adjective and
noun vectors.  They hypothesized that the length of the vectors thus
constructed can be used to distinguish acceptable and semantically
deviant adjective-noun composites.  Using a few hundred adjective-noun
composites selected by humans for evaluation, they found that deviant
composites have a shorter vector than acceptable ones, in accordance
with their expectation.  In contrast to their work, our approach does
not require extensive human evaluation.

Recent theoretical work \cite{Arora2015} has approached the problem of
explaining the so-called ``compositionality'' property exhibited by some
word embeddings.  In that work, unnormalised vectors are used in their
model of the word relation task.  It is hoped that experimental
approaches such as those described here might enable theoretical
investigations to describe the role of the word vector length in the
word relation tasks.
\end{section}

\begin{section}{Corpus and model}\label{corpus-and-model}
Our training data is built from the Wikipedia data dump from October
2013.  To remove the bulk of robot-generated pages from the training
data, only pages with at least 20 monthly page views are
retained.\footnote{For further justification and to obtain the dataset,
  see\\ \blogpost} Stubs and disambiguation pages are also removed,
leaving 463 thousand pages with a total of 482 million words.
Punctuation marks and numbers were removed from the pages and all words
were lower-cased.  Word frequencies are summarised in
Table~\ref{word-occurrences-table}.  This base corpus is then modified
as described in Sections~\ref{WFVE} and \ref{CNVE}.  For
recognisability, the tokens inserted into the corpus are upper-cased.

\begin{subsection}{Word2vec}\label{word2vec}
Word2vec, a feed-forward neural network with a single hidden layer,
learns word vectors from word co-occurrences in an unsupervised manner.
Word2vec comes in two versions.  In the continuous bag-of-words (CBOW)
model, the words appearing around a target word serve as input.  That
input is projected linearly onto the hidden layer and the network then
attempts to predict the target word on output.  Training is achieved
through back-propagation.  The word vectors are encoded in the weights
of the first synaptic layer, ``syn0''.  The weights of the second
synaptic layer (``syn1neg'', in the case of negative sampling) are
typically discarded.  In the other model, called skip-gram, target and
context words swap places, so that the target word now serves as input,
while the network attempts to predict the context words on output.

@@

The geometric distribution is truncated to limit the number of tokens
inserted into the corpus.  For any choice $0 < p < 1$ and maximum value
$n > 0$, the truncated geometric distribution is given by the
probability density function
\begin{equation} 
\label{distro}
P_{p, n} (i) = \frac{p^{i-1}  (1-p)}{1 - p^n}, \qquad 1
\leqslant i \leqslant n.
\end{equation} 
The factor in the denominator, which tends to unity in the limit $n \to
\infty$, assures proper normalisation.  We have chosen this distribution
because the probabilities decay exponentially base $p$ as a function of
$i$.  Of course, other distributions might equally well have been chosen
for the experiments.
\end{subsection}
\end{section}

@@

\begin{subsection}{Tokens derived from existing words}\label{WFVEexisting}
We choose uniformly at random a small number of words from the
unmodified vocabulary for our experiment.  In order that the inserted
tokens do not have too low a frequency, only words which occur at least
10 thousand times are chosen.  We also include the high-frequency
stopword \word{the} for comparison.  Table~\ref{word-frequency-counts}
lists the words chosen for this experiment along with their frequencies.

The replacement procedure of Section~\ref{replacement-procedure} is then
performed for each of these words, using a geometric decay rate of $p =
\tfrac{1}{2}$, and maximum value $n=20$, so that the 1st token is
inserted with a probability of about $0.5$, the 2nd with a probability of
about $0.25$, and so on.  This value of $p$ is one of a range of values
that ensure that, for each word, multiple tokens will be inserted with a
frequency sufficient to survive the low-frequency cut-off of
128.  A maximum value $n=20$ suffices for this choice of $p$, since
$2^{20 + \log_2{128}}$ exceeds the maximum frequency of any token in
the corpus.  Figure~\ref{fig:word-frequency-experiment-text-cat}
illustrates the effect of these modifications on a sample text, with a
family of tokens \word{CAT\_i}, derived from the word \word{cat}.
Notice that all occurrences of the word \word{cat} have been replaced
with the tokens \word{CAT\_i}.

\begin{table}
	\input{word-frequency-experiment-counts.tex}
	\caption{Words chosen for the word frequency experiment, along with their frequency in the unmodified corpus. }
	\label{word-frequency-counts}
\end{table}

\begin{figure}
	\begin{mdframed}
	\input{word-frequency-experiment-text-cat.tex}
	\end{mdframed}
	\caption{Example sentences modified in the word frequency
          experiment as per Section~\ref{WFVEexisting}, where the word
          \word{cat} is replaced with tokens \word{CAT\_i} using the
          truncated geometric distribution (\ref{distro}) with
          $p=\tfrac{1}{2}$ and $n=20$.}
	\label{fig:word-frequency-experiment-text-cat}
\end{figure}

@@

\begin{subsection}{Tokens derived from an artificial, meaningless word}\label{WFVEmeaningless}
Whereas the tokens introduced above all replace an existing word that
carries a meaning, we now include for comparison a high-frequency,
meaningless word.  We choose to introduce an artificial, entirely
meaningless word \word{VOID} into the corpus, rather than choose an
existing (stop)word whose meaninglessness is only supposed.  To achieve
this, we intersperse the word uniformly at random throughout the corpus so
that its relative frequency is $0.005$.  The co-occurrence
distribution of \word{VOID} thus coincides with the unconditional word
distribution.  The replacement procedure is then performed for this
word, using the same values for $p$ and $n$ as above.
Figure~\ref{fig:word-frequency-experiment-text-void} shows the effect of
these modifications on a sample text, where a higher relative frequency of $0.05$ is used instead
for illustrative purposes.  Although the probability of inserting token
\word{VOID\_1} is twice that of inserting token \word{VOID\_2} and four
times that of inserting token \word{VOID\_3}, these three tokens all
appear once in this small piece of text.

\begin{figure}
	\begin{mdframed}
	\input{word-frequency-experiment-text-void.tex}
	\end{mdframed}
	\caption{The same example sentences as in
          Figure~\ref{fig:word-frequency-experiment-text-cat} where
          instead of the word \word{cat} now the meaningless word
          \word{VOID} is replaced with tokens \word{VOID\_i}.  For
          illustrative purposes, the meaningless word \word{VOID} was
          here interspersed with a relative frequency of $0.05$.}
	\label{fig:word-frequency-experiment-text-void}
\end{figure}
\end{subsection}

@@

\subsubsection{Word frequency and vector direction}\label{WFVE-direction}
Figure~\ref{word-frequency-experiment-heatmap} shows the cosine
similarity of pairs of vectors representing some of the tokens used in
this experiment.  Recall that the cosine similarity measures the extent
to which two vectors have the same direction, taking a maximum value of
$1$ and a minimum value of $-1$.  The number of different tokens
associated with an experiment word is the number of times that its
frequency can be halved and remains above the low-frequency cut-off of
$128$.

Consider first the words other than \word{the} and \word{VOID}.  These
words typically occur in a limited number of distinct contexts, and so a
limited number of samples is required to capture their co-occurrence
distributions.  We therefore expect that the vectors representing the
tokens derived from any of these words can be properly learned from the
modified corpus, even though their combined frequency equals the
frequency of that word in the original corpus.  The red blocks in
Figure~\ref{word-frequency-experiment-heatmap} clearly demonstrate that
the direction of these vectors is constant for each family of tokens.
In other words, the direction of these vectors is only determined by the
original word and is independent of how often that word appears,
provided sufficient samples have been processed for the word vector to
settle in its equilibrium direction.

The artificial word \word{VOID}, on the other hand, occurs by
construction in every context.  A much higher number of samples is
therefore required to capture its co-occurrence distribution, and
thereby to learn its word vector.  The same is true, but to a lesser
extent, for the stopword \word{the}.  The difficulty of learning the
vector representing \word{VOID\_i} or \word{THE\_i} increases with $i$,
for the tokens occur increasingly less often.  This is consistent with
Figure~\ref{word-frequency-experiment-heatmap}, where it is apparent
that the vectors representing \word{VOID\_i} for small $i$, say $i = 1,
\dots, 5$, roughly share a common direction, while this is no longer
true for higher values of $i$.  The same trend is apparent, although
less extremely so, for the word \word{the}.  The vectors for the tokens
\word{THE\_i} share a common direction for values of $i$ as high as
$14$, before they start changing direction.  We conclude that these
deviations are artifacts of poor learning due to an insufficient number
of occurrences of these tokens.
Table~\ref{word-frequency-experiment-cosine-similarity}, showing the
most similar words to each token \word{THE\_i}, supports this further.
In the discussions below, the tokens \word{VOID\_i} and \word{THE\_i}
whose vectors have a cosine similarity less than $0.8$ with
\word{VOID\_1} and \word{THE\_1}, respectively, are therefore discarded.

@@

\subsubsection{Word frequency and vector length}
We next consider the effect of frequency on word vector length.
Throughout, we measure vector length using the Euclidean norm.
Figure~\ref{fig:word-frequency-experiment-graph} shows this relation for
individual words, both for the word vectors, represented by the weights
of the first synaptic layer, syn0, in the word2vec neural network, and
for the vectors represented by the weights of the second synaptic layer,
syn1neg.  We include the latter, which are typically ignored, for
completeness.  Each line corresponds to a single word, and the points on
each line indicate the frequency and vector length of the tokens derived
from that word.  For example, the six points on the line corresponding
to the word \word{protestant} are labeled, from right to left, by the
tokens \word{PROTESTANT\_1}, \word{PROTESTANT\_2}, \dots,
\word{PROTESTANT\_6}.  Again, the number of points on the line is
determined by the frequency of the original word.  For example, the
frequency of the word \word{protestant} can be halved at most $6$ times
so that the frequency of the last token is still above the
low-frequency cut-off.  Because all the points on a line share the same
co-occurrence distribution, the left panel in
Figure~\ref{fig:word-frequency-experiment-graph} demonstrates conclusively
that length does indeed depend on frequency directly.  Moreover, this
relation is seen to be approximately linear for each word considered.
Notice also that the relative positions of the lengths of the word vectors associated
with the experiment words are roughly independent of the frequency band,
i.e., the plotted lines rarely cross.

Notice the inflection of the syn1neg curve in
Figure~\ref{fig:word-frequency-experiment-graph} for \word{the} at
approximately $10^6$.  We have no explanation of this.  Also observe
that the lengths of the vectors representing the meaningless tokens
\word{VOID\_i} are approximately constant (about $2.5$).  Since we already found the
direction to be also constant, it is sensible to speak of the word
vector of \word{VOID} irrespective of its frequency.  In particular, the
token \word{VOID\_1} may be taken as an approximation.

@@

\begin{sidewaysfigure*}
	\centering{\includegraphics[scale=0.6]{word-frequency-experiment-graph}}
	\caption{ Vector length vs.\ frequency for tokens derived from a
          few words chosen at random.  For each word, tokens of varying
          frequency but with the co-occurrence distribution of that word
          were inserted into the corpus, as described in
          Section~\ref{WFVE}.  The vectors are obtained from the first
          synaptic layer, syn0, of the word2vec neural network.  The
          vectors obtained from the second layer, syn1neg, are included
          for completeness.  Legend entries are ordered by vector length
          of the left-most data point in the syn0 plot, descending.}
	\label{fig:word-frequency-experiment-graph}
\end{sidewaysfigure*}
\end{section}

\begin{section}{Varying co-occurrence noise}\label{CNVE}
This second experiment is complementary to the first.  Whereas in the
first experiment we studied the effect of word frequency on word vectors
for fixed co-occurrence, we here study the effect of co-occurrence noise
when the frequency is fixed.  As before, we do so in a controlled
manner.

\subsection{Generating noise}
We take the noise distribution to be the (observed) unconditional word
distribution.  Thus noise can be added to the co-occurrence distribution
of a word by simply interspersing occurrences of that word uniformly at
random throughout the corpus.  A word that is consistently used in a
distinctive context in the unmodified corpus thus appears in the
modified corpus also in completely unrelated contexts.  As in
Section~\ref{WFVE}, we choose a small number of words from the
unmodified corpus for this experiment.
Table~\ref{cooccurrence-noise-words} lists the words chosen, along with
their frequencies in the corpus.

For each of these words, the replacement procedure of
Section~\ref{replacement-procedure} is performed using a geometric rate
of decay of $p = \tfrac{5}{6}$, and truncating the distribution at
$n=8$.  For every replacement token (e.g. \word{CAT\_i}), occurrences of
this token are subsequently interspersed uniformly at random throughout
the corpus, such that the frequency of the replacement token is restored
to that of the original word \word{cat}.  For example, if the original
word \word{cat} occurred $1000$ times, then after the replacement
procedure, \word{CAT\_2} occurs approximately $694$ times, so a further
(approximately) $306$ random occurrences of \word{CAT\_2} are
interspersed throughout the corpus.  Thus the word \word{cat} is removed
from the corpus, and a family of tokens \word{CAT\_i}, $1 \leqslant i
\leqslant 8$ is introduced.  These tokens all have the same frequency as
the word \word{cat}, but their co-occurrence distributions, while based
on that of \word{cat}, have an increasing amount of noise.

Figure~\ref{fig:co-occurrence-noise-experiment-text} illustrates the
effect of this modification in the case where the only word chosen is
\word{cat}.  The original text in this case concerned both cats and
dogs.  Notice that the word \word{cat} has been replaced entirely in the
cats section by \word{CAT\_i} and, moreover, that these same tokens
appear also in the dogs section. These occurrences (and additionally,
with probability, some occurrences from the cats section) constitute
noise.

The proportion of occurrences of the token \word{CAT\_i} that arise by
replacing the word \word{cat} is $P_{5/6, 8}(i)$, and the proportion
that arise from randomly interspersing this token throughout the corpus
is $r_i := 1 - P_{5/6, 8}(i)$.  The particular value of $p$ (large
compared to that of Section~\ref{WFVE}) was chosen so that the
proportion of noise $r_i$ did not increase too quickly with $i$.  These
proportions are the sampling points on our axis of variation, and this
value of $p$ (together with the choice of $n$) assures a reasonable
spread.  Other parameter values (or indeed other distributions) could,
of course, have been used equally well.
